Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-04 Manuscript/manuscript.Rmd
addressing kellen comments

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained for the autosomal chromosomes [?? Don't think sentence is necessary ??]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution [?? Don't think this sentence is needed ??].

@@

Using the sets of cell line specific genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries. They are described in detail below:

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a bin overlapped with regions assigned to genomic annotations (features) was calculated. In the case of no overlaps, the value was set to 0.

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0. [?? This entire section feels extremely wordy. Probably can reduce to a couple sentences ??]

@@

The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison.

@@

## Cell line and Data Resolution Specific Model Construction

>>>>

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2). Functional annotations were represented by chromosome-specific regions on the linear genome obtained by ChIP-seq peak calling algorithms.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Using the sets of cell line specific genomic annotations, we created three types of predictors that were used to describe the spatial relationship with genomic bins, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. They are described in detail below:

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a bin overlapped with regions assigned to genomic annotations (features) was calculated.

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin.

@@

The cell line- and resolution-specific data, composed of the responsed vector and predictor-type specific data matrix  then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison.

@@

## CELL LINE AND DATA RESOLUTION SPECIFIC MODEL CONSTRUCTION
